---
output:
    bookdown::html_document2: default
---
# Understanding Distributions {#Distributions}

```{r, library-theme, include=FALSE}
library(tidyverse)
theme_set(theme_light())
```

Every distribution discovered in the probability world is determined by a **random variable**. These variables are the not the same ones that you were exposed to in algebra or calculus class. Instead, they are used to map random processes, like rolling dice or playing the lottery, to readable notation.

We'll begin with a simple example of this by flipping a coin. Let the random variable X (usually random variables are capital letters) denote this process. It can take two forms, heads or tails. We can use a binary 0 or 1 to show this:

$$X = \{0,1\}$$

You might be asking yourself: "What does this have to do with probability?" We begin by defining the random variable so that we can later discover its mathematical attributes, such as its mean or variance, for example.

We can also define the probability that the random variable $X$ (the coin) takes the form of heads or tails by using probability notation.

$$P(X = x)$$

Where x can equal 0 or 1. We can interpret the statement above as "the probability that the random variable X equals x." So what is that probability? By common sense, we can conclude that the probability that you get heads is $P(X=1)=1/2$. We can extend this logic to show that **the sum of probabilities for any random variable is 1**.

$$P(X = 0) + P(X = 1)= 1/2 + 1/2 = 1$$

In other, more complicated probability distributions, we use random variables such as $X$ to mark these processes without having to explain them each time.

## Uniform Distribution {#Uniform}

We'll begin our journey with uniform distribution. Given $a < b$, we can define this distribution's **density** as:

$$ P(X=x) = \frac{1}{b-a}\quad \textrm{for} \enspace a \leq x \leq b $$

When we refer to densities we are talking about probabilities. The two terms are interchangeable, but the term **density** emphasizes the finite probability space that a distribution occupies. As mentioned previously, the sum of probabilities for any distribution is 1.

How do we get this equation? Imagine drawing a rectangle in the interval (a,b) with height 1/(b-a) as in Figure \@ref(fig:uniform-basic).

```{r, uniform-basic, fig.cap= "Remember Scales? See if you can understand the code", fig.align="center", out.width="80%"}
ggplot() +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1), 
            fill = NA,
            color = "black") +
  labs(x = "x", y = "Density") +
  scale_x_continuous(breaks = c(0,1), labels = c("a", "b")) +
  scale_y_continuous(breaks = c(0,1), labels = c("", "h"))
```

We have the height as $\frac{1}{(b-a)}$. We know this since the cumulative area of a distribution must be 1 and the width is b-a. Using the area of a rectangle with h being the height of the uniform distribution.

$$(b-a)h=1$$

For the equation to hold, h must equal 1/(b-a).

Let's define X as a **discrete** random variable that is uniform on 1 through 10. The term "discrete" means that $X$ can only be integers, and the fact that its probabilities are uniform means that every integer has an equal probability of occurring. Therefore:

$$
\begin{split}
&X=\{1,2,...,10\}\\&P(X=x)=\frac{1}{10-1}=1/9\quad\textrm{for}\quad1\leq x\leq 10 
\end{split}
$$

The first line outlines the random variable X and the numbers that it can take. The second line then calculates the probability of the random variable taking the integers 1-10 as 1/9. In other words, the **density**, or individual probability, for each of $1\leq x\leq 10$ is 1/9.

The first `stats` function we can use with the uniform distribution is `dunif` which also returns the density at point x in the distribution given left and right bounds. This is quite easy to determine in the uniform distribution because **every point has the same density of h**- in other words, the height. `dunif` has 3 arguments.

1. A vector of values to calculate densities for.

2. The minimum or `min` of the distribution.

3. The maximum or `max` of the distribution.

We'll return to our example where the beginning and end of the uniform distribution is at x = 1 and x = 10, respectively. What are the densities at x = 1, 5, and 10?

```{r, uniform-densities}
dunif(c(1, 5, 10), min = 1, max = 10)
```

You are correct if you guessed 1/9 or approximately 0.111! If $a = 1$, $b = 10$ then by definition $h = 1/9$. Thus, as we calculated earlier, every point in the distribution has a density of 1/9.

We can visualize this by drawing a rectangle using `geom_rect` and then adding points with `geom_point`. Remember that certain geometries require aesthetics beyond x and y. 

```{r, geomrect-plot, fig.cap= "Uniform distribution with marked densities", fig.align="center", out.width="80%"}
ggplot() +
  geom_rect(aes(xmin = 1, xmax = 10, ymin = 0, ymax = 1), fill = NA, 
            color = "black") +
  labs(x = "x", y = "Density") +
  geom_point(aes(x = c(1, 5, 10), y = 1), size = 5, color = "navy")
```

The `punif()` function helps us find the cumulative density between at the qth quantile. Using a different rectangle with $a = 0$, $b = 4$, and $h = 1/4$, we can find the proportion of the distribution between the minimum of 0 and maximum of 2 by writing the following:

```{r, punif}
punif(2, 0, 4, lower.tail = TRUE) # Remember: we don't always have to name our arguments if we know the order.
```

We can similarly write:

```{r, punif-alternative}
sum(dunif(0:1, min = 0, max = 4)) # Not including 2.
```

There are three arguments in this function: a vector of quantiles, a minimum, maximum, and a binary `lower.tail` argument. By default this is set to `TRUE`.

We can also think about the `punif()` function as drawing a smaller rectangle from 0 to 2 (if `lower.tail = FALSE`) and calculating its area. Below, this is the same as the percent of the total area that the navy rectangle occupies.

```{r, punif-plot, fig.cap= "Uniform distribution with cumulative density marked", fig.align="center", out.width="80%"}
ggplot() +
  geom_rect(aes(xmin = 0, xmax = 2, ymin = 0, ymax = 1/4),  # Small rectangle
            alpha = .2, 
            fill = "navy", 
            color = "navy", 
            linetype = "dashed") +
    geom_rect(aes(xmin = 0, xmax = 4, ymin = 0, ymax = 1/4), # Large rectangle
              fill = NA, 
              color = "black") +
  labs(x = "x", y = "Density")
```

$$ area \hspace{0.2 cm}= \hspace{0.2 cm}base \hspace{0.2cm} * \hspace{0.2 cm}height$$

$$area = 2*0.25=0.5$$

Now we will consider the opposite scenario where we want to find the x-value that corresponds with the 50th percentile of our distribution. This is the purpose of `qunif()`. We already know from using the distribution function `punif()` that this is 2. 

```{r, punif-opposite}
qunif(0.5, 0, 4) #lower.tail is also set to TRUE by default
```

The last function `runif()` generates random deviates within the distribution. There are three arguments in this function:

- n, the number of deviates we want to produce

- min, the left bound of the uniform distribution

- max, the right bound of the uniform distribution

We'll use the `round()` function to round them to the hundredths place.

```{r, uniform-deviates, fig.cap= "Uniform distribution with random deviates", fig.align="center", out.width="80%"}
# Create object of deviates

unif_dev <- round(runif(10, min = 0, max = 4), digits = 2)

# Plot

ggplot() +
  geom_rect(aes(xmin = 0, xmax = 4, ymin = 0, ymax = 1/4), 
            fill = NA,
            color = "black") +
  geom_point(aes(x = unif_dev, y = 1/4), size = 5, color = "navy") + # Plot deviates
  labs(x = "x", y = "Density") +
  scale_x_continuous(breaks = seq(0, 4, by = 1), labels = seq(0, 4, by = 1))
```

The fact that the deviates appeared on non-integer values (i.e. those that would normally be represented by fractions or decimals) reveals that the uniform distribution is not always discrete. In this case, as well as others later in this guide, distributions are labeled **continuous**. As we investigate several other types of distributions, we'll start with the discrete cases and then transition to the continuous ones.
