---
output:
    bookdown::html_document2: default
---
# Basic Concepts of ggplot {#ggplot}

A visualization can be anything. It can be a drawing, a photograph, a sculpture, a well-decorated cake... you get the idea. However, data scientists and statisticians are often not very skilled at all of these art forms, so they are limited to their tools at hand: functions, code, and geometries.

This leads us to data visualizations. What separates a data visualization from a drawing is the tools used to construct it. In this guide, our blank canvas is a function called `ggplot`.

```{r, empty-plot, message= FALSE, warning = FALSE, fig.cap="An empty plot", out.width="80%", fig.asp=.75, fig.align="center"}
library(tidyverse)
ggplot()
```

## Geometries and Aesthetic Mappings

Every time that we want to visualize something, we must use this function. The stuff that fills up this space are called **geometries**. Here are just a few that ggplot has available to us.

- `geom_rect` produces rectangles.
- `geom_point` creates dots.
- `geom_line` draws lines.
- `geom_bar` makes a barplot.
- `geom_function` uses functions to draw a continuous curve.

If the geometries are the shapes on our ggplot canvas, how do we put them on there? This is where aesthetics and mapping come in. When we call the function `ggplot` we need to connect our data to aesthetic mappings which are then applied to various geometries.

But where do we use aesthetics mappings? The `mapping` argument in the `ggplot` and `geom` functions are what connect variables from our dataset. Lets look at a dataset of Marvel superheroes and apply the following aesthetics:

```{r, avengers-formatting, include=FALSE}
library(fivethirtyeight)
data(avengers)

glimpse(avengers)

avengers <- avengers %>%
  select(name_alias, gender, appearances, years_since_joining, death1) %>% 
  arrange(desc(appearances)) %>%
  rename(death = death1) %>% 
  slice(1:20) 

avengers$gender = str_to_lower(avengers$gender)

avengers$gender[avengers$name_alias == "Susan Richards (nee Storm)"] <- "female"
```

Years since joining Marvel → x

Appearances → y 

Gender → color

Now we will use the `aes` function to define our mappings for a simple scatterplot. We will also use the `labs` function to label our axes and title and an additional `size` argument to make our points larger.

```{r, avengers-plot, fig.align="center", out.width="80%", fig.cap = "Basic Avengers scatterplot"}
ggplot(data = avengers, mapping = aes(x = years_since_joining,
                                      y = appearances,
                                      color = gender)) +
  geom_point(size = 5) +
  labs(title = "'Age' of Superheroes and Appearances",
       x = "Years since Joining Marvel",
       y = "Appearances")
```

We used two arguments in the ggplot function: data and mapping. Next, we added the geometry `geom_point` using a plus sign (+). The aesthetic mapping in the ggplot function are then passed onto `geom_point`.

Even though we must include data and mappings to produce a visualization, we do not always have to name the arguments themselves. For instance, we could simply write:

```{r, avengers-simple, eval=F, results = "hide"}
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     color = gender)) +
  geom_point()
```

Aesthetic mappings can also produce different geometries by simply changing the function. For example, instead of using `geom_point` we can use `geom_rect` to draw rectangles or squares.

```{r, avengersplot-rect, fig.align="center", out.width="80%", fig.cap="Avengers scatterplot with rectangles"}
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     fill = gender)) +
  geom_rect(aes(xmin = years_since_joining - 1,
                xmax = years_since_joining + 1,
                ymin = appearances - 100,
                ymax = appearances + 100)) +
    labs(title = "'Age' of Superheroes and Appearances",
       x = "Years since Joining Marvel",
       y = "Appearances")
```

Notice in Figure \@ref(fig:avengersplot-rect) that the `geom_rect` geometry requires aesthetic mappings beyond those in `geom_point`, including the minimums and maximums for x and y. The color aesthetic of `geom_point` turns to fill in `geom_rect` since the rectangles are not points; they are shapes with empty space.

So far we have established that aesthetics can be passed onto different geometries. 

Figure \@ref(fig:ggplot-diagram) recaps where we are at now.

```{r, ggplot-diagram, results="markup", echo=FALSE, fig.align="center", out.width="80%", fig.cap="ggplot diagram"}
DiagrammeR::grViz("digraph {

graph [layout = dot, rankdir = LR]

# define the global styles of the nodes. We can override these in box if we wish
node [shape = rectangle, style = filled, fillcolor = Linen]

data [label = 'Data']
ggplot [label = 'ggplot']
map [label =  'Mappings \n Aesthetics']
geometries [label = 'Geometries']
plot [label= 'Plot']

# edge definitions with the node IDs
{data map}  -> ggplot -> geometries -> plot
}")
```

Generally, we want to provide mappings to the ggplot function so that they pass to all of our geometries. Including them in every line of code would be repetitive.

```{r, avengers-aes, eval=F, results = "hide"}
# Don't do this!
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     color = gender)) +
  geom_point(aes(x = years_since_joining,
                 y = appearances,
                 color = gender))

# Do this instead!
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     color = gender)) +
  geom_point()
```

But there are plenty of instances in which we wouldn't want to keep our aesthetics limited to the ggplot function line--for example, plots with two different groups or those with different scales on the y axis (although it is not generally recommended to have multiple scales on an axis since it is confusing). Let's return to the Avengers plot but keep the `color` argument in `geom_point` so that the color of our text doesn't differ by gender.

```{r, avengersplot-gender, fig.align="center", out.width="80%", fig.cap = "Avengers scatterplot with unique color and label aesthetic assignments"}
ggplot(avengers, aes(x = years_since_joining, y = appearances)) +
  geom_point(aes(color = gender), size = 5) +
  geom_text(aes(label = name_alias), size = 3) +
  labs(title = "'Age' of Superheroes and Appearances",
       x = "Years since Joining Marvel",
       y = "Appearances")
```

Figure \@ref(fig:avengersplot-gender) doesn't look pretty but you get the idea. A general rule to follow with data visualizations is to avoid mapping one aesthetic to multiple variables. For instance, a plot in which the color is used to illustrate both an Avenger's gender and their name would be confusing.

```{r, avengersplot-wrong, fig.align="center", out.width="80%", fig.cap= "Incorrect Aesthetic Mapping"}
# Don't do this!
ggplot(avengers, aes(x = years_since_joining, y = appearances)) +
  geom_point(aes(color = gender), size = 5) +
  geom_text(aes(color = name_alias, label = name_alias), size = 3) +
  labs(title = "'Age' of Superheroes and Appearances",
       x = "Years since Joining Marvel",
       y = "Appearances")
```

Sadly, Steve Rogers is not a gender and neither are the other Avenger names in the legend. This is why we should not map one aesthetic to more than one variable.

## Altering Aesthetics and Scalings 

Now, let's introduce a way to change our aesthetics further. There are some arguments within geom functions that allow us to do this. For instance, you might have noticed that I used `size` to make points and labels larger in the previous plots. We can also use `color` or `fill` to make adjustments to our geoms.

```{r, avengersplot-red, fig.align="center", out.width="80%", fig.cap= "Avengers scatterplot in red!"}
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances)) +
  geom_point(color = "tomato", size = 5)
```

But previously we have related color with an Avenger's gender. ggplot uses some default colors to differentiate male and female Avengers but we can change this further with **scales**. Every scale function has three components:

1. `scale_`
2. The name of the aesthetic we are scaling. In this case, `color`.
3. A method of applying the scale. In this case, we'll use `manual`.

So to combine all of the previous steps together, we would call the function`scale_color_manual` to directly alter the colors of our aesthetic related to gender.

```{r, avengersplot-scale, fig.align="center", out.width="80%", fig.cap = "Avengers scatterplot using `scale_color_manual`"}
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     color = gender)) +
  geom_point(size = 5) +
  scale_color_manual(values = c("Tomato", "Navy"))
```

We've not only learned about the gender gap in the Marvel universe but also how to apply scales to our aesthetics! There are a number of scales that we can use further. For example, `scale_alpha_binned` allows us to alter the transparency of points after providing a `range` between 0-1. The `breaks` argument separates our observations into different groups.

```{r, avengersplot-alpha, fig.align="center", out.width="80%", fig.cap = "Avengers scatterplot with `scale_alpha_continuous`"}
ggplot(avengers, aes(x = years_since_joining,
                     y = appearances,
                     alpha = appearances)) +
  geom_point(size = 5) +
  scale_alpha_continuous(range = c(0.2, 1),
                         breaks = c(2000, 3000, 4000))
```

An important fact to remember is that mapping a variable to more than one aesthetic sometimes exaggerate differences in data. In this case, using two aesthetics--y and alpha--for appearance count draws the eye to the Avengers with the most appearances. This is why any data scientist must consider the [principles of data visualization](https://socviz.co/index.html#preface) and how certain techniques mislead or distort data.

So far, we have learned how to call the `ggplot` function, apply aesthetic mappings, layer geometries, and alter aesthetics with scales. Any data visualization, regardless of its complexity, follows this formula. 

Hopefully this chapter has provided you a some more confidence as you begin producing your own visualizations. However, this is by no means an exhaustive tutorial. To learn more about data visualizations here are some useful resources:

- [R Cookbook](https://www.amazon.com/gp/product/0596809158/ref=as_li_tf_tl?ie=UTF8&camp=1789&creative=9325&creativeASIN=0596809158&linkCode=as2&tag=cooforr09-20)
- [Cookbook for R](http://www.cookbook-r.com/)
- [The R Graph Gallery](https://www.r-graph-gallery.com/index.html) 
- [R for Data Science Online Community](https://www.rfordatasci.com/)

The rest of this guide specifically uses our new visualization tool to learn about the statistical distributions that govern our world. Click on to learn more!
